A quantum parallel Markov chain Monte Carlo

### 2022-07-28 main.tex
about to start short revision
more writing on fixed point

Uncountably many single- and two-qubit quantum gates exist, but the real power of quantum computing stems from the development of complex quantum gates that act on multiple qubits simultaneously.   To access this power, we need one more tool that also appears in the statistics literature.  The \emph{Kronecker} or \emph{tensor} product between an $L$-by-$M$ matrix $\A$ and an $N$-by-$O$ matrix $\B$ is the $LN$-by-$MO$ matrix
\begin{align}\label{eq:kron}
\A \otimes \B = \begin{bmatrix}
A_{11} \B & \dots & A_{1M} \B \\
\vdots & \ddots & \vdots \\
A_{L1}\B & \cdots & A_{LM}  \B
\end{bmatrix} \, .
\end{align} 
In statistics, the Kronecker product features in the definition of a matrix normal distribution and is sometimes helpful when specifying the kernel function of a multivariate Gaussian process \citep{werner2008estimation}. Here, the product is equivalent to the parallel action of individual quantum gates on individual qubits.  Simple application of Formula \eqref{eq:kron} shows that
\begin{align}\label{eq:twoqubits}
\H \otimes \H = \frac{1}{2} \left[\begin{array}{rrrr}
1 & 1 & 1 & 1 \\
1 &  -1& 1 & -1 \\
1 & 1 & -1 & -1 \\
1 & -1 & -1 & 1
\end{array}\right] \quad \mbox{and} \quad \ket{0}\otimes\ket{0}=\begin{bmatrix}
1 \\ 0
\end{bmatrix} \otimes \begin{bmatrix}
1 \\ 0
\end{bmatrix} = \begin{bmatrix}
1 \\ 0 \\ 0 \\ 0
\end{bmatrix} \, .
\end{align} 
We may also denote the left product $\H^{\otimes2}$ and the right product $\ket{0}^{\otimes2}$, $\ket{00}$ or $\ket{0}\ket{0}$. One may therefore express Equation \eqref{eq:quantparallel} as a series of transformations applied to the 4-vector on the very right. Letting $\ket{10}$, $\ket{01}$ and $\ket{11}$ take on analogous meanings to $\ket{00}$, an immediate result of Equation \eqref{eq:twoqubits} is that
\begin{align}\label{eq:twodmult}
\H^{\otimes2} \ket{00} = \frac{1}{2}\Big(\ket{00} + \ket{01} + \ket{10} + \ket{11}\Big) \, .
\end{align}
The action of $\H^{\otimes2}$ transforms $\ket{00}$ into a superposition of the states $\ket{00}$, $\ket{10}$, $\ket{01}$ and $\ket{11}$, where the probability of selecting each is a uniform $(1/2)^2=1/4$. Writing so many $0$s and $1$s is tedious, so an alternative notation becomes preferable. Exchanging $\ket{0}$ for $\ket{00}$, $\ket{1}$ for $\ket{01}$, $\ket{2}$ for $\ket{10}$, and $\ket{3}$ for $\ket{11}$, Equation \eqref{eq:twodmult} becomes the more succinct
\begin{align*}
\H^{\otimes2} \ket{00} = \frac{1}{2}\Big(\ket{0} + \ket{1} + \ket{2} + \ket{3}\Big) \, .
\end{align*}
This formula extends generally to operations over $n$ qubits.  Now letting $N=2^n$, we have
\begin{align*}%\label{eq:superpos}
\H^{\otimes n} \ket{0}^{\otimes n} = \frac{1}{\sqrt{N}}\Big(\ket{0} + \ket{1} + \dots + \ket{N-1}\Big) =  \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x}=:\ket{h} \, ,
\end{align*}
and we call $\ket{h}$ a \emph{uniform superposition} over the states $\ket{0},\dots,\ket{N-1}$. The many-qubit analogue for the quantum parallelism of Equation \eqref{eq:quantparallel} is then
\begin{align*}%\label{eq:quantparallel2}
\ket{0}^{\otimes n}\ket{0} \longrightarrow \left(\frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \right) \ket{0}  \longrightarrow \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \ket{f(x)} \, .
\end{align*}
In the next section, we review established quantum algorithms that leverage such many-qubit quantum parallelism for search and minimization.  Following this, we embed parallel MCMC subroutines within quantum minimization.

@@

 While Grover's algorithm delivers impressive speedups over classical search algorithms, it has a major weakness.
 Figure \ref{fig:grov} hides the fact that the probability of success for Grover's algorithm is \emph{not} monatonic in the number of iterations.  Running the algorithm for more than $R$ iterations can backfire.  For example, running the algorithm for $\sqrt{2N}$ iterations when $M=1$ results in a probability of success less than $0.095$ \citep{boyer1998tight}.  The non-monotinicity of Grover's algorithm becomes particularly problematic when we do not know the number of solutions $M$. Taking for example $N=2^{20}$, \citet{boyer1998tight} point out that 804 iterations provide an extremely high probability of success when $M=1$ but a one-in-a-million chance of success when $M=4$.  To solve this problem and develop an effective search algorithm when $M$ is unknown, those authors adopt the strategy of focusing on the expected number of iterations before success. In particular, they propose the quantum exponential search algorithm (Algorithm \ref{alg:expoSearch}).  When a solution exists, the algorithm returns a solution with expected total number of Grover iterations bounded above by $\frac{9}{2}\sqrt{N/M}$.  Still better, this upper bound reduces to $\frac{9}{4}\sqrt{N/M}$ for the special case $M\ll N$, and simulations presented in Figure \ref{fig:expo} show that even this bound is large.  Such results come in handy when deciding whether to halt the algorithm's progress if one believes it possible that no solutions exist.  Indeed, this turns out to be useful in the context of quantum minimization.

 \begin{figure}[!t]
 	\centering
 	\includegraphics[width=0.9\linewidth]{expoSearch.pdf}
 	\caption{Total number of oracle evaluations required by quantum exponential searching algorithm \citep{boyer1998tight} for different numbers of solutions $M$ and search set sizes $N$ from 500 independent simulations each.  Horizontal lines at $\frac{9}{4}\sqrt{N}$ represent upper bounds on expected total number of evaluations to obtain a solution for the $M=1$ problem.}\label{fig:expo}
 \end{figure}

@@

>>>>

@@

 While Grover's algorithm delivers impressive speedups over classical search algorithms, it has a major weakness.
 Figure \ref{fig:grov} hides the fact that the probability of success for Grover's algorithm is \emph{not} monatonic in the number of iterations.  Running the algorithm for more than $R$ iterations can backfire.  For example, running the algorithm for $\sqrt{2N}$ iterations when $M=1$ results in a probability of success less than $0.095$ \citep{boyer1998tight}.  The non-monotinicity of Grover's algorithm becomes particularly problematic when we do not know the number of solutions $M$. Taking for example $N=2^{20}$, \citet{boyer1998tight} point out that 804 iterations provide an extremely high probability of success when $M=1$ but a one-in-a-million chance of success when $M=4$.  To solve this problem and develop an effective search algorithm when $M$ is unknown, those authors adopt the strategy of focusing on the expected number of iterations before success. In particular, they propose the quantum exponential search algorithm (Algorithm \ref{alg:expoSearch}).  When a solution exists, the algorithm returns a solution with expected total number of Grover iterations bounded above by $\frac{9}{2}\sqrt{N/M}$.  Still better, this upper bound reduces to $\frac{9}{4}\sqrt{N/M}$ for the special case $M\ll N$, and simulations presented in Figure \ref{fig:expo} show that even this bound is large.  Such results come in handy when deciding whether to halt the algorithm's progress if one believes it possible that no solutions exist.  Indeed, this turns out to be useful in the context of quantum minimization (Section \ref{sec:qmin}).

 \begin{figure}[!t]
 	\centering
 	\includegraphics[width=1\linewidth]{searchingFigs.pdf}
 	\caption{[Left] Total number of oracle evaluations required by quantum exponential searching algorithm (QESA) \citep{boyer1998tight} for different numbers of solutions $M$ and search set sizes $N$ from 500 independent simulations each.  Horizontal lines at $\frac{9}{4}\sqrt{N}$ represent upper bounds on expected total number of evaluations to obtain a solution for the $M=1$ problem. [Right] Under the same experimental settings, additional number of oracle evaluations required by fixed-point quantum search (FPQS) algorithm of \citet{yoder2014fixed} compared to QESA.}\label{fig:expo}
 \end{figure}

Algorithm \ref{alg:expoSearch} is not the only quantum search algorithm that is useful when the number of solutions $M$ is unknown.  \citet{yoder2014fixed} use \emph{generalized Grover iterations} that extend Equation \eqref{eq:grov} to include general phases $(\alpha_j,\beta_j)$:
\begin{align}\label{eq:grovGen}
	\G(\alpha_j,\beta_j) := - \U_h(\alpha_j) \U_{x_0}(\beta_j) := \Big((1-e^{-i\alpha_j})\ket{h}\bra{h}-\I \Big) \Big( \I - (1-e^{i\beta_j}) \ket{x_0}\bra{x_0} \Big) \, .
\end{align}
The method first fixes an error tolerance $\delta^2>0$ and a lower bound $w\leq \lambda=M/N$ and then selects an odd upper bound $L$ on the total number of oracle queries ($L-1$) such that 
\begin{align*}
	L \geq \frac{\log(2/\delta)}{\sqrt{w}} \, .
\end{align*}
Finally,  letting $l=(L-1)/2$, one obtains phases $(\alpha_j,\beta_j)$ for $j=1,\dots,l$ that satisfy
\begin{align*}
	\alpha_j = - \beta_{l-j+1} =2 \cot^{-1} \left( \tan(2\pi j /L) \sqrt{1-\gamma^2} \right) \, ,
\end{align*}
 where $\gamma^{-1} =  \cos(\frac{1}{L}\cos^{-1}(\frac{1}{\delta}))$ is the reciprocal of the inverse  $L$th Chebyshev polynomial of the first kind.

@@

Uncountably many single- and two-qubit quantum gates exist, but the real power of quantum computing stems from the development of complex quantum gates that act on multiple qubits simultaneously.   To access this power, we need one more tool that also appears in the statistics literature.  The \emph{Kronecker} or \emph{tensor} product between an $L$-by-$M$ matrix $\A$ and an $N$-by-$O$ matrix $\B$ is the $LN$-by-$MO$ matrix
\begin{align}\label{eq:kron}
	\A \otimes \B = \begin{bmatrix}
		A_{11} \B & \dots & A_{1M} \B \\
		\vdots & \ddots & \vdots \\
		A_{L1}\B & \cdots & A_{LM}  \B
	\end{bmatrix} \, .
\end{align} 
In statistics, the Kronecker product features in the definition of a matrix normal distribution and is sometimes helpful when specifying the kernel function of a multivariate Gaussian process \citep{werner2008estimation}. Here, the product is equivalent to the parallel action of individual quantum gates on individual qubits.  Simple application of Formula \eqref{eq:kron} shows that
\begin{align}\label{eq:twoqubits}
	\H^{\otimes 2} =\H \otimes \H = \frac{1}{2} \left[\begin{array}{rrrr}
		1 & 1 & 1 & 1 \\
		1 &  -1& 1 & -1 \\
		1 & 1 & -1 & -1 \\
		1 & -1 & -1 & 1
	\end{array}\right] \quad \mbox{and} \quad \ket{0}\otimes\ket{0}=\begin{bmatrix}
		1 \\ 0
	\end{bmatrix} \otimes \begin{bmatrix}
		1 \\ 0
	\end{bmatrix} = \begin{bmatrix}
		1 \\ 0 \\ 0 \\ 0
	\end{bmatrix} \, .
\end{align} 
We may also denote the left product $\H^{\otimes2}$ and the right product $\ket{0}^{\otimes2}$, $\ket{00}$ or $\ket{0}\ket{0}$. One may therefore express Equation \eqref{eq:quantparallel} as a series of transformations applied to the 4-vector on the very right. Letting $\ket{10}$, $\ket{01}$ and $\ket{11}$ take on analogous meanings to $\ket{00}$, an immediate result of Equation \eqref{eq:twoqubits} is that
\begin{align}\label{eq:twodmult}
	\H^{\otimes2} \ket{00} = \frac{1}{2}\Big(\ket{00} + \ket{01} + \ket{10} + \ket{11}\Big) \, .
\end{align}
The action of $\H^{\otimes2}$ transforms $\ket{00}$ into a superposition of the states $\ket{00}$, $\ket{10}$, $\ket{01}$ and $\ket{11}$, where the probability of selecting each is a uniform $(1/2)^2=1/4$. Writing so many $0$s and $1$s is tedious, so an alternative notation becomes preferable. Exchanging $\ket{0}$ for $\ket{00}$, $\ket{1}$ for $\ket{01}$, $\ket{2}$ for $\ket{10}$, and $\ket{3}$ for $\ket{11}$, Equation \eqref{eq:twodmult} becomes the more succinct
\begin{align*}
	\H^{\otimes2} \ket{00} = \frac{1}{2}\Big(\ket{0} + \ket{1} + \ket{2} + \ket{3}\Big) \, .
\end{align*}
This formula extends generally to operations over $n$ qubits.  Now letting $N=2^n$, we have
\begin{align*}%\label{eq:superpos}
	\H^{\otimes n} \ket{0}^{\otimes n} = \frac{1}{\sqrt{N}}\Big(\ket{0} + \ket{1} + \dots + \ket{N-1}\Big) =  \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x}=:\ket{h} \, ,
\end{align*}
and we call $\ket{h}$ a \emph{uniform superposition} over the states $\ket{0},\dots,\ket{N-1}$. The many-qubit analogue for the quantum parallelism of Equation \eqref{eq:quantparallel} is then
\begin{align*}%\label{eq:quantparallel2}
	\ket{0}^{\otimes n}\ket{0} \longrightarrow \left(\frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \right) \ket{0}  \longrightarrow \frac{1}{\sqrt{N}} \sum_{x=0}^{N-1} \ket{x} \ket{f(x)} \, .
\end{align*}
